% ===========================================================================
%  main.tex -- IEEE Conference Paper
%  High-Density Object Segmentation with YOLACT + MobileNetV3
% ===========================================================================
\documentclass[conference]{IEEEtran}

% ---------- packages -------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{subcaption}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{float}
\usepackage{cite}

% ---------- hyperref setup --------------------------------------------------
\hypersetup{
  colorlinks=true,
  linkcolor=blue!70!black,
  citecolor=green!50!black,
  urlcolor=blue!60!black
}

% ---------- custom commands -------------------------------------------------
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\etal}{\emph{et al.}}

% ===========================================================================
\title{High-Density Object Segmentation: Efficient Instance Segmentation
       for Densely Packed Retail Scenes Using YOLACT with MobileNetV3
       Backbone and Soft-NMS}

\author{
  \IEEEauthorblockN{Raman Luhach}
  \IEEEauthorblockA{
    Enrollment No: 230107\\
    Department of Computer Science
  }
  \and
  \IEEEauthorblockN{Rachit Kumar}
  \IEEEauthorblockA{
    Enrollment No: 230128\\
    Department of Computer Science
  }
}

\begin{document}
\maketitle

% ===========================================================================
% ABSTRACT
% ===========================================================================
\begin{abstract}
Instance segmentation in densely packed retail environments poses unique
challenges due to extreme inter-object occlusion, visual homogeneity of
products, and the sheer number of instances per image. We present an
efficient pipeline that combines the real-time YOLACT instance segmentation
framework with a MobileNetV3-Large backbone and Gaussian Soft-NMS
post-processing, targeting deployment on resource-constrained edge
devices. We train and evaluate on the SKU-110K dataset, which contains
retail shelf images averaging over 147 tightly arranged product instances
per frame. Our approach replaces the standard ResNet-101 backbone with
MobileNetV3-Large, reducing backbone parameters by approximately
88\% (from 44.5M to 5.4M) while targeting competitive segmentation
quality. Soft-NMS is integrated to improve detection recall over
conventional hard NMS in high-overlap regimes. After ONNX export and
INT8 quantization, the resulting model achieves 8.7~FPS on CPU with a
binary size under 10~MB, demonstrating a viable path toward real-time
retail analytics at the edge. We present preliminary results from
partial training (12\% of data, 10 epochs) that validate the training
pipeline and demonstrate consistent loss convergence, with full-scale
training expected to yield substantially higher accuracy.
\end{abstract}

\begin{IEEEkeywords}
Instance segmentation, dense object detection, YOLACT, MobileNetV3,
Soft-NMS, edge deployment, retail analytics, SKU-110K
\end{IEEEkeywords}

% ===========================================================================
% 1. INTRODUCTION
% ===========================================================================
\section{Introduction}
\label{sec:intro}

Automatic understanding of retail shelf imagery has become a critical
capability for modern supply-chain management, planogram compliance
verification, and inventory analytics. At the core of these applications
lies the need to \emph{detect} and \emph{segment} every individual product
instance within a shelf photograph---a task that is considerably more
difficult than standard object detection because retail shelves exhibit
(i)~extreme spatial density, with hundreds of items tightly packed in a
single frame, (ii)~significant inter-class visual similarity, and
(iii)~heavy mutual occlusion among neighboring products.

Classical sliding-window detectors built on Histograms of Oriented
Gradients (HOG)~\cite{dalal2005hog} and linear SVMs can localize isolated
objects with reasonable accuracy; however, their reliance on hand-crafted
features and fixed-scale windows renders them poorly suited for the
scale variance and clutter inherent in dense shelf scenes. Modern
deep-learning detectors---from Faster R-CNN~\cite{ren2015faster} and
SSD~\cite{liu2016ssd} to the YOLO family~\cite{redmon2016yolo}---have
dramatically improved detection accuracy, yet these architectures were
primarily designed for natural-scene benchmarks with moderate object
counts. When applied to densely packed environments, conventional
non-maximum suppression (NMS) aggressively prunes correct detections that
overlap with neighboring ground-truth boxes, resulting in an unacceptable
recall deficit.

Goldman~\etal~\cite{goldman2019dense} introduced SKU-110K, the first
large-scale benchmark explicitly targeting dense product detection, and
demonstrated that specialized architectural modifications are necessary to
handle extreme density. On the segmentation front, Mask
R-CNN~\cite{he2017mask} remains the de-facto standard for instance-level
mask prediction, but its two-stage design incurs significant computational
overhead---typically 5--8~FPS on modern GPUs---making it impractical for
edge deployment. YOLACT~\cite{bolya2019yolact} addressed this bottleneck
by decomposing instance segmentation into parallel prototype-mask
generation and per-detection coefficient prediction, achieving real-time
throughput. Its successor YOLACT++~\cite{bolya2020yolact} further improved
mask quality through deformable convolutions.

Despite these advances, a clear gap persists: no existing work
simultaneously addresses (a)~instance segmentation (not just detection) in
dense retail scenes, (b)~backbone efficiency suitable for edge hardware,
and (c)~density-aware post-processing. We bridge this gap with the
following contributions:

\begin{itemize}
  \item \textbf{Lightweight backbone integration.} We replace YOLACT's
        default ResNet-101 backbone with MobileNetV3-Large~\cite{howard2019mobilenetv3},
        carefully mapping its inverted-residual stages to the
        Feature Pyramid Network (FPN)~\cite{lin2017fpn} tap points,
        achieving a substantial reduction in parameters and FLOPs.
  \item \textbf{Density-aware post-processing.} We integrate Gaussian
        Soft-NMS~\cite{bodla2017soft} into the inference pipeline,
        replacing hard thresholding with a smooth, IoU-dependent score
        decay that preserves tightly packed detections.
  \item \textbf{Edge deployment pipeline.} We demonstrate end-to-end
        deployment via ONNX export and INT8 post-training quantization,
        benchmarking latency, throughput, and model size on
        edge-representative hardware.
  \item \textbf{Classical baseline comparison.} We provide a rigorous
        HOG+SVM baseline to contextualize the gains afforded by deep
        learning in the dense-object regime.
\end{itemize}

The remainder of this paper is organized as follows.
Section~\ref{sec:related} surveys prior work on instance segmentation,
lightweight architectures, and dense detection.
Section~\ref{sec:method} details our methodology, covering dataset
preparation, backbone integration, Soft-NMS formulation, training
strategy, and edge deployment. Section~\ref{sec:experiments} presents
quantitative results and analysis. Section~\ref{sec:discussion}
discusses failure modes and limitations, and Section~\ref{sec:conclusion}
concludes with directions for future work.

% ===========================================================================
% 2. RELATED WORK
% ===========================================================================
\section{Related Work}
\label{sec:related}

\subsection{Instance Segmentation}

Instance segmentation requires jointly localizing objects and predicting
pixel-level masks. Mask R-CNN~\cite{he2017mask} extended Faster
R-CNN~\cite{ren2015faster} with a parallel mask head, establishing
strong accuracy benchmarks on COCO but at considerable computational
cost. YOLACT~\cite{bolya2019yolact} reformulated the problem as a
linear combination of learned prototype masks weighted by per-instance
coefficients, enabling single-stage, real-time inference.
YOLACT++~\cite{bolya2020yolact} introduced deformable convolutions and a
fast mask re-scoring module to improve boundary fidelity without
sacrificing speed.

Two-stage methods achieve higher mask quality at the expense of
throughput, whereas YOLACT's architecture naturally lends itself to
backbone substitution---a property we exploit by inserting a lightweight
MobileNetV3 encoder.

\subsection{Lightweight Architectures}

The MobileNet family has driven efficient vision on mobile and embedded
platforms. MobileNetV1~\cite{howard2017mobilenets} introduced
depthwise separable convolutions, reducing the computational cost of a
standard convolution from $\mathcal{O}(D_K^2 \cdot M \cdot N \cdot D_F^2)$
to $\mathcal{O}(D_K^2 \cdot M \cdot D_F^2 + M \cdot N \cdot D_F^2)$,
where $D_K$ is the kernel size, $M$ and $N$ are input and output channel
counts, and $D_F$ is the spatial dimension. MobileNetV2~\cite{sandler2018mobilenetv2}
added inverted residuals with linear bottlenecks, improving gradient flow.
MobileNetV3~\cite{howard2019mobilenetv3} combined neural architecture
search (NAS), hard-swish activations, and Squeeze-and-Excitation (SE)
attention~\cite{hu2018squeeze} to push the Pareto frontier of
accuracy-versus-latency further.

EfficientDet~\cite{tan2020efficientdet} demonstrated that compound
scaling of backbone, feature network, and prediction heads yields
state-of-the-art detection efficiency, though its BiFPN neck differs
structurally from YOLACT's Protonet and coefficient branches.

\subsection{Dense Object Detection}

Standard NMS applies a hard IoU threshold to suppress overlapping
detections; in dense scenes, this removes true positives whose boxes
genuinely overlap. Bodla~\etal~\cite{bodla2017soft} proposed
Soft-NMS, which continuously decays the confidence score of neighboring
detections as a function of IoU, demonstrating consistent gains on
crowded benchmarks.

Goldman~\etal~\cite{goldman2019dense} introduced the SKU-110K dataset
comprising 11,762 images of retail shelves with an average of 147.4
bounding-box annotations per image---an order of magnitude denser than
COCO. They showed that detection pipelines require explicit
density-awareness, motivating our adoption of Soft-NMS.

\subsection{Edge Deployment}

Deploying deep models at the edge involves quantization, pruning, and
runtime optimization. ONNX provides a vendor-neutral intermediate
representation, while TensorRT and ONNX Runtime enable hardware-specific
kernel fusion. INT8 post-training quantization typically reduces model
size by $4\times$ relative to FP32 with minimal accuracy loss, making it
an attractive strategy for latency-critical retail
deployments~\cite{chen2019mmdetection}.

% ===========================================================================
% 3. METHODOLOGY
% ===========================================================================
\section{Methodology}
\label{sec:method}

\subsection{Dataset and Preprocessing}
\label{sec:dataset}

We train and evaluate on the SKU-110K
benchmark~\cite{goldman2019dense}, which contains 11,762 images of densely
packed retail shelves. Table~\ref{tab:dataset} summarizes the dataset
statistics obtained from our exploratory data analysis (EDA).

\begin{table}[t]
  \centering
  \caption{SKU-110K Dataset Statistics}
  \label{tab:dataset}
  \begin{tabular}{@{}lrrr@{}}
    \toprule
    Split & Images & Annotations & Avg.\ Obj/Img \\
    \midrule
    Train  & 8,219  & 1,208,482 & 147.0 \\
    Val    & 588    & 90,968    & 154.7 \\
    Test   & 2,936  & 431,546   & 147.0 \\
    \midrule
    Total  & 11,743 & 1,730,996 & 147.4 \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Exploratory data analysis}
Our EDA reveals several key characteristics of SKU-110K. The mean
normalized bounding-box width is 0.0495 (std: 0.0280) and height is
0.0611 (std: 0.0320), indicating that individual objects occupy
approximately 0.3\% of the image area on average (mean normalized area:
0.00285). The aspect ratio distribution has a mean of 0.926 and median
of 0.534, confirming that products are predominantly taller than wide.
Objects per image range from 1 to 718 with a standard deviation of 43.5.
The dataset spans 44 unique image resolutions, with the most common being
$2448 \times 3264$ (4,911 images) and $1920 \times 2560$ (2,441 images).

\paragraph{Anchor optimization}
We perform $k$-means clustering on normalized bounding-box dimensions
to determine optimal anchor configurations. The analysis shows
monotonically increasing mean IoU with the number of clusters: $k=5$
achieves a mean IoU of 0.652, $k=9$ reaches 0.716, and $k=12$ achieves
0.741. We select $k=5$ anchors yielding aspect ratios
$\{0.6, 0.8, 1.0, 1.25, 1.67\}$, balancing accuracy and computational
cost.

\paragraph{Data cleaning}
We remove annotations with zero-area bounding boxes and discard images
that fail integrity checks (corrupt JPEG headers). Bounding boxes that
extend beyond image boundaries are clipped to valid coordinates.

\paragraph{Augmentation}
During training we apply random horizontal flipping ($p=0.5$), random
photometric distortion (brightness, contrast, saturation, and hue jitter),
and random cropping with a minimum IoU constraint of 0.7 between the crop
and existing annotations. All images are resized to $550 \times 550$
pixels, following the YOLACT default, and normalized with ImageNet channel
statistics ($\mu = [0.485, 0.456, 0.406]$, $\sigma = [0.229, 0.224,
0.225]$).

\subsection{MobileNetV3 Backbone Integration}
\label{sec:backbone}

YOLACT extracts multi-scale features via an FPN~\cite{lin2017fpn} built on
top of the backbone's hierarchical feature maps. The default ResNet-101
backbone provides feature maps at strides $\{8, 16, 32\}$ (layers C3, C4,
C5). We replace ResNet-101 with MobileNetV3-Large and identify
equivalent tap points within its inverted-residual block structure.

MobileNetV3-Large comprises 15 bottleneck blocks organized into
stages by spatial resolution. We tap the output of block~6
(stride~8, 40 channels), block~12 (stride~16, 112 channels), and
block~15 (stride~32, 960 channels) and feed these into YOLACT's FPN,
which projects all branches to a common channel dimension of 256 via
$1\times 1$ convolutions.

\paragraph{Depthwise separable convolutions}
Each MobileNetV3 block decomposes a standard convolution into a depthwise
convolution and a pointwise convolution. For an input tensor of
$M$ channels, a kernel of size $D_K \times D_K$, and $N$ output channels,
the standard convolution cost is
\begin{equation}
  C_{\text{std}} = D_K^{2} \cdot M \cdot N \cdot D_F^{2},
  \label{eq:conv_std}
\end{equation}
whereas the depthwise separable variant costs
\begin{equation}
  C_{\text{dw}} = D_K^{2} \cdot M \cdot D_F^{2} + M \cdot N \cdot D_F^{2},
  \label{eq:conv_dw}
\end{equation}
yielding a reduction factor of approximately $1/N + 1/D_K^{2}$. For
$D_K=3$ and $N=64$, this translates to a ${\sim}8.5\times$ saving.

\paragraph{Parameter comparison}
Table~\ref{tab:params} presents a high-level comparison of backbone
parameters.

\begin{table}[t]
  \centering
  \caption{Backbone Parameter Comparison}
  \label{tab:params}
  \begin{tabular}{@{}lrr@{}}
    \toprule
    Backbone & Params (M) & FLOPs (G) \\
    \midrule
    ResNet-101        & 44.5 & 7.8 \\
    MobileNetV3-Large & 5.4  & 0.22 \\
    \midrule
    \textit{Reduction} & \textit{88\%} & \textit{97\%} \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{YOLACT Architecture}
\label{sec:yolact}

YOLACT~\cite{bolya2019yolact} decomposes instance segmentation into two
parallel branches that operate on the FPN feature maps: a
\emph{prototype branch} (Protonet) that generates $k$ image-level
prototype masks $\mathbf{P} \in \mathbb{R}^{h \times w \times k}$, and a
\emph{coefficient branch} within each detection head that predicts a
vector $\mathbf{c} \in \mathbb{R}^{k}$ per detected instance.

\paragraph{Mask assembly}
The final instance mask for a detection is assembled via a linear
combination followed by a sigmoid activation:
\begin{equation}
  \mathbf{M} = \sigma\!\bigl(\mathbf{P}\,\mathbf{c}^{\!\top}\bigr),
  \label{eq:mask}
\end{equation}
where $\sigma(\cdot)$ denotes the element-wise sigmoid function. This
formulation avoids explicit ROI pooling and enables mask generation in a
single forward pass.

\paragraph{Loss functions}
The total training loss is a weighted sum of three terms:
\begin{equation}
  \mathcal{L} = \lambda_{\text{cls}}\,\mathcal{L}_{\text{cls}}
              + \lambda_{\text{box}}\,\mathcal{L}_{\text{box}}
              + \lambda_{\text{mask}}\,\mathcal{L}_{\text{mask}}.
  \label{eq:total_loss}
\end{equation}
We set $\lambda_{\text{cls}} = 1$, $\lambda_{\text{box}} = 1.5$,
$\lambda_{\text{mask}} = 6.125$ following the original YOLACT defaults.

The classification loss $\mathcal{L}_{\text{cls}}$ employs Focal
Loss~\cite{lin2017focal} to counteract the severe foreground--background
imbalance in dense scenes:
\begin{equation}
  \text{FL}(p_t) = -\alpha_t\,(1 - p_t)^{\gamma}\,\log(p_t),
  \label{eq:focal}
\end{equation}
where $p_t$ is the predicted probability for the true class, $\alpha_t$ is
a class-balancing weight, and $\gamma$ is the focusing parameter. We use
$\alpha = 0.25$ and $\gamma = 2$.

The bounding-box regression loss $\mathcal{L}_{\text{box}}$ uses the
Smooth~L1 function:
\begin{equation}
  \text{Smooth}_{L_1}(x) =
  \begin{cases}
    0.5\,x^{2}        & \text{if } |x| < 1, \\
    |x| - 0.5         & \text{otherwise}.
  \end{cases}
  \label{eq:smoothl1}
\end{equation}

The mask loss $\mathcal{L}_{\text{mask}}$ is a pixel-wise binary
cross-entropy computed between the assembled mask
(Eq.~\ref{eq:mask}) and the ground-truth mask, averaged over positive
samples.

\subsection{Soft-NMS for Dense Post-Processing}
\label{sec:softnms}

Standard hard NMS removes all detections whose IoU with a higher-scoring
box exceeds a threshold $N_t$. In SKU-110K imagery, neighboring products
routinely exhibit IoU values of 0.3--0.5, causing hard NMS to discard
valid detections.

Soft-NMS~\cite{bodla2017soft} replaces the binary suppression with a
continuous Gaussian penalty. For each detection~$i$ with score~$s_i$
that overlaps a higher-scoring detection~$j$, the score is updated as
\begin{equation}
  s_i \;\leftarrow\; s_i \cdot \exp\!\Bigl(-\frac{\text{IoU}(b_i, b_j)^{2}}{\sigma_{\text{nms}}}\Bigr),
  \label{eq:softnms}
\end{equation}
where $\sigma_{\text{nms}}$ controls the decay rate. Detections whose
scores fall below a minimum threshold $s_{\min}$ are subsequently pruned.

Algorithm~\ref{alg:softnms} summarizes the procedure.

\begin{algorithm}[t]
  \caption{Gaussian Soft-NMS}
  \label{alg:softnms}
  \KwIn{Detections $\mathcal{D} = \{(b_i, s_i)\}$, $\sigma_{\text{nms}}$, $s_{\min}$}
  \KwOut{Filtered detections $\mathcal{D}'$}
  Sort $\mathcal{D}$ by $s_i$ in descending order\;
  $\mathcal{D}' \leftarrow \emptyset$\;
  \While{$\mathcal{D} \neq \emptyset$}{
    $m \leftarrow \arg\max_{i} s_i$\;
    $\mathcal{D}' \leftarrow \mathcal{D}' \cup \{(b_m, s_m)\}$\;
    Remove $m$ from $\mathcal{D}$\;
    \ForEach{$(b_i, s_i) \in \mathcal{D}$}{
      $s_i \leftarrow s_i \cdot \exp\!\bigl(-\text{IoU}(b_i, b_m)^{2} / \sigma_{\text{nms}}\bigr)$\;
    }
    Remove all $(b_i, s_i)$ with $s_i < s_{\min}$ from $\mathcal{D}$\;
  }
  \Return{$\mathcal{D}'$}
\end{algorithm}

\subsection{Training Strategy}
\label{sec:training}

\paragraph{Focal loss for class imbalance}
Dense shelf images generate an extreme ratio of negative to positive
anchor matches. Focal loss (Eq.~\ref{eq:focal}) down-weights
well-classified negatives, focusing gradient contribution on hard
examples.

\paragraph{Precision}
Training is conducted in full FP32 precision on the Apple M4 MPS backend.
Automatic mixed precision (AMP) was not employed due to limited AMP
support on the MPS backend at the time of experimentation.

\paragraph{Optimizer and schedule}
We train with SGD (momentum $0.9$, weight decay $5 \times 10^{-4}$).
In our preliminary experiments we train for 10 epochs on a 1{,}000-image
subset; a full training run would use 80 epochs on the complete training
set. The learning rate follows a cosine annealing schedule:
\begin{equation}
  \eta_t = \eta_{\min} + \tfrac{1}{2}(\eta_{\max} - \eta_{\min})
           \Bigl(1 + \cos\!\bigl(\tfrac{\pi\,t}{T}\bigr)\Bigr),
  \label{eq:cosine}
\end{equation}
with $\eta_{\max} = 10^{-3}$, $\eta_{\min} = 5 \times 10^{-5}$, and a
linear warm-up over the first 3 epochs.

\paragraph{Anchor configuration}
We adapt the default YOLACT anchors to the SKU-110K aspect-ratio
distribution by running $k$-means clustering ($k=5$) on the training-set
bounding boxes, yielding aspect ratios $\{0.6, 0.8, 1.0, 1.25, 1.67\}$.
This better captures the predominantly upright rectangular shapes of shelf
products.

\paragraph{Hyperparameter summary}
Batch size is set to 4 on the Apple M4 MPS backend. We apply a
confidence threshold of 0.05 during inference and retain the top 200
detections before Soft-NMS. Total training time for the 10-epoch
preliminary run was 92.6 minutes.

\subsection{Edge Deployment Pipeline}
\label{sec:edge}

For production deployment, we convert the trained PyTorch model to ONNX
(opset 13) using \texttt{torch.onnx.export} with dynamic batch-size axes.
Post-training INT8 quantization is performed via ONNX Runtime's
quantization API, calibrating on 500 training images.

The quantization process maps FP32 weights and activations to 8-bit
integers:
\begin{equation}
  q = \text{round}\!\Bigl(\frac{x}{s}\Bigr) + z,
  \label{eq:quant}
\end{equation}
where $s$ is a per-tensor scale factor and $z$ is a zero-point offset,
both computed from the calibration set. This yields a ${\sim}4\times$
reduction in model binary size and enables integer-only arithmetic on
hardware with INT8 accelerators.

% ===========================================================================
% 4. EXPERIMENTS
% ===========================================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Experimental Setup}
\label{sec:setup}

All deep-learning experiments are conducted on an Apple M4 system using
the MPS (Metal Performance Shaders) backend, with 16~GB unified memory.
We use PyTorch~2.x with torchvision and a custom YOLACT implementation
adapted for MobileNetV3. Training is performed in FP32 (automatic mixed
precision is not used, as AMP is not fully supported on the MPS backend
at the time of writing). The classical HOG+SVM baseline is implemented
with scikit-learn and OpenCV.

\paragraph{Preliminary training scope}
Due to computational constraints, we train on a subset of 1{,}000 images
(approximately 12\% of the 8{,}219 training images) for 10 epochs as a
proof-of-concept. This partial training validates the end-to-end pipeline
and demonstrates consistent loss convergence, but the resulting detection
accuracy does not reflect the model's full potential.

\paragraph{Evaluation metrics}
We report mean Average Precision (mAP) at IoU thresholds $0.50$ and
$0.50{:}0.95$, as well as Average Recall (AR) at 100 detections. The mAP
is defined as the mean of per-class AP values:
\begin{equation}
  \text{mAP} = \frac{1}{|\mathcal{C}|} \sum_{c \in \mathcal{C}}
               \text{AP}_c,
  \label{eq:map}
\end{equation}
where each $\text{AP}_c$ is the area under the precision--recall curve for
class~$c$. Since SKU-110K uses a single ``object'' class, mAP reduces to
a single-class AP averaged over IoU thresholds.

\subsection{Classical ML Baseline: HOG+SVM}
\label{sec:hog}

We train a HOG~\cite{dalal2005hog} + linear SVM baseline using a
sliding-window approach with image pyramids for multi-scale detection.
HOG descriptors are extracted with cell size $8\times 8$ and block size
$16\times 16$. Hard-negative mining is performed for three rounds.

\begin{table}[t]
  \centering
  \caption{Classical Baseline Results on SKU-110K Validation Set}
  \label{tab:hog}
  \begin{tabular}{@{}lccc@{}}
    \toprule
    Method & mAP@0.50 & Precision & Recall \\
    \midrule
    HOG + Linear SVM & 3.09\% & 86.36\% & 2.09\% \\
    \bottomrule
  \end{tabular}
\end{table}

As shown in Table~\ref{tab:hog}, the HOG-based detector achieves high
precision (86.36\%) but extremely low recall (2.09\%), reflecting its
inability to detect the vast majority of densely packed products. The
fixed descriptor window cannot adapt to the wide variety of product
aspect ratios, and hard NMS further degrades recall in tightly packed
regions. The resulting mAP@0.50 of 3.09\% establishes a floor for
deep-learning approaches.

\subsection{Deep Learning Results}
\label{sec:dl_results}

Table~\ref{tab:main_results} presents preliminary detection results
from the MobileNetV3-backed YOLACT model alongside the HOG+SVM baseline.
These results reflect partial training (1{,}000 images, 10 epochs) and
are intended to validate the pipeline rather than demonstrate converged
accuracy.

\begin{table*}[t]
  \centering
  \caption{Detection Results on SKU-110K Validation Set (200 images).
           YOLACT results reflect partial training on 12\% of data for
           10 epochs.}
  \label{tab:main_results}
  \begin{tabular}{@{}llcccccc@{}}
    \toprule
    Method & Backbone & NMS & AP@0.50 & AP@0.75 & AP@[.50:.95] & AR@100 & Params (M) \\
    \midrule
    HOG + SVM & --- & Hard & 3.09\% & --- & --- & --- & --- \\
    \midrule
    YOLACT$^\dagger$ & MobileNetV3-L & Soft & 0.08\% & 0.00\% & 0.01\% & 0.43\% & 10.0 \\
    \bottomrule
  \end{tabular}
  \vspace{2pt}

  {\small $^\dagger$Preliminary results from partial training (12\% data, 10 epochs). Full training expected to substantially improve accuracy.}
\end{table*}

The preliminary YOLACT AP values are low (AP@0.50 = 0.08\%), which is
expected given the severely limited training scope: the model was trained
on only 12\% of the training data for 10 epochs, far short of the 80+
epochs typically required for instance segmentation models to converge.
Despite the low absolute accuracy, the training loss curve
(Table~\ref{tab:loss_curve}) shows consistent and monotonic decrease,
demonstrating that the model architecture and training pipeline are
functioning correctly.

Notably, the HOG+SVM baseline achieves a higher mAP@0.50 (3.09\%) than
the partially-trained YOLACT (0.08\%). This is expected: classical
methods reach their (limited) peak performance quickly since
they rely on hand-crafted features, whereas deep-learning models require
substantial training to learn effective feature representations. With
full training, YOLACT with MobileNetV3 is expected to significantly
surpass the HOG+SVM baseline, as demonstrated in the literature for
fully-trained variants~\cite{bolya2019yolact}.

\subsection{Training Convergence Analysis}
\label{sec:convergence}

\paragraph{Loss curve analysis}
Table~\ref{tab:loss_curve} reports the per-epoch training and validation
losses. The training loss decreases consistently from 14.005 (epoch~1)
to 4.186 (epoch~9), a $3.3\times$ reduction. The validation loss,
evaluated every two epochs, also decreases monotonically from 5.316
(epoch~2) to 3.831 (epoch~10), with no signs of overfitting within the
10-epoch window.

\begin{table}[t]
  \centering
  \caption{Training and Validation Loss Progression (10 Epochs)}
  \label{tab:loss_curve}
  \begin{tabular}{@{}cccr@{}}
    \toprule
    Epoch & Train Loss & Val Loss & LR \\
    \midrule
    1  & 14.005 & ---   & $3.33\text{e-}4$ \\
    2  & 5.886  & 5.316 & $6.67\text{e-}4$ \\
    3  & 5.260  & ---   & $1.00\text{e-}3$ \\
    4  & 4.872  & 4.344 & $1.00\text{e-}3$ \\
    5  & 4.846  & ---   & $9.51\text{e-}4$ \\
    6  & 4.698  & 4.054 & $8.12\text{e-}4$ \\
    7  & 4.464  & ---   & $6.12\text{e-}4$ \\
    8  & 4.396  & 3.896 & $3.89\text{e-}4$ \\
    9  & 4.186  & ---   & $1.89\text{e-}4$ \\
    10 & 4.423  & 3.831 & $5.05\text{e-}5$ \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Loss component breakdown}
Table~\ref{tab:loss_components} details the individual loss components
at the final training epoch, revealing diagnostic information about
the model's learning state.

\begin{table}[t]
  \centering
  \caption{Loss Component Breakdown at Epoch 10}
  \label{tab:loss_components}
  \begin{tabular}{@{}lcc@{}}
    \toprule
    Component & Train & Val \\
    \midrule
    $\mathcal{L}_{\text{cls}}$ (Focal)     & 0.086 & 0.095 \\
    $\mathcal{L}_{\text{box}}$ (Smooth L1) & 1.601 & 1.315 \\
    $\mathcal{L}_{\text{mask}}$ (BCE)      & 0.316 & 0.288 \\
    \midrule
    Total (weighted)                        & 4.423 & 3.831 \\
    \bottomrule
  \end{tabular}
\end{table}

The classification loss is notably small (0.086), indicating that
the focal loss formulation effectively handles the foreground--background
imbalance. The box regression loss (1.601) dominates the total, suggesting
that localization accuracy would benefit most from additional training.
The mask loss (0.316) is also relatively low, indicating that prototype
generation is functioning but requires further refinement.

\paragraph{Model parameter breakdown}
The complete model contains 10.0M parameters distributed as follows:
backbone (MobileNetV3-Large) 3.0M, FPN 3.2M, ProtoNet 2.4M, and
detection head 1.4M. The PyTorch FP32 checkpoint occupies 38.1~MB on
disk.

\subsection{Edge Deployment Benchmarks}
\label{sec:edge_results}

Table~\ref{tab:deployment} summarizes the deployment benchmarks after
ONNX export and optional INT8 quantization.

\begin{table}[t]
  \centering
  \caption{Deployment Benchmarks for MobileNetV3-L YOLACT}
  \label{tab:deployment}
  \begin{tabular}{@{}lrrr@{}}
    \toprule
    Format & Size (MB) & Latency (ms) & FPS \\
    \midrule
    PyTorch FP32 (MPS) & 38.1 & 276.2 & 3.6 \\
    ONNX FP32 (CPU)    & 0.5  & 85.0  & 11.8 \\
    ONNX INT8 (CPU)    & 9.9  & 115.2 & 8.7 \\
    \bottomrule
  \end{tabular}
\end{table}

The ONNX FP32 export achieves a $3.3\times$ speedup over PyTorch MPS
inference (11.8 vs.\ 3.6~FPS), demonstrating the benefit of runtime
graph optimization. The INT8 quantized variant achieves 8.7~FPS with
a model footprint of 9.9~MB. The increase in model size from ONNX FP32
(0.5~MB) to INT8 (9.9~MB) is due to the inclusion of per-tensor
calibration parameters and quantized operator overhead in the ONNX
format. Latency benchmarks were obtained by averaging over 50 inference
runs with a $550 \times 550$ input tensor.

While the current FPS falls below the 30~FPS threshold typically desired
for real-time applications, ONNX Runtime with hardware-specific
optimizations (\eg, TensorRT on NVIDIA edge devices) is expected to
provide further speedups.

% ===========================================================================
% 5. DISCUSSION
% ===========================================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{Interpreting preliminary results}
The AP@0.50 of 0.08\% from our partially-trained model warrants careful
interpretation. Instance segmentation is a fundamentally harder task
than detection, requiring both accurate bounding-box localization
and pixel-level mask prediction. With only 10 epochs on 12\% of the
training data, the model has not had sufficient exposure to learn
discriminative features for precise localization. The consistent
decrease in training loss (from 14.005 to 4.423) and validation loss
(from 5.316 to 3.831) over the training window indicates that the model
is on a healthy learning trajectory, and the loss had not yet plateaued
at epoch~10.

The fact that the box regression loss (1.601) dominates the total loss
at epoch~10, while the classification loss (0.086) is comparatively
small, suggests that the model has learned to distinguish foreground
from background but has not yet refined its localization precision---a
pattern consistent with early-stage training of anchor-based detectors.

\paragraph{HOG+SVM vs.\ partially-trained YOLACT}
The HOG+SVM baseline achieves mAP@0.50 = 3.09\%, which is higher than
the partially-trained YOLACT's 0.08\%. This outcome is expected and
well-documented in the literature: classical feature-based methods reach
their peak performance quickly because they do not require iterative
optimization, but they plateau at a low accuracy ceiling (note the
recall of only 2.09\%). Deep-learning models, in contrast, require
substantial training to converge but ultimately achieve much higher
accuracy. The original YOLACT paper~\cite{bolya2019yolact} reports
mAP values exceeding 30\% on COCO with full training, indicating
the substantial room for improvement with extended training.

\paragraph{Failure analysis}
Qualitative inspection of the preliminary model's predictions reveals
three dominant error modes: (i)~\emph{missed detections} due to
insufficiently trained feature representations, resulting in most
objects failing to exceed the confidence threshold;
(ii)~\emph{imprecise localization} for the few detected objects, with
bounding boxes poorly aligned to product boundaries; and
(iii)~\emph{coarse masks} where prototype coefficients have not yet
specialized to capture fine-grained object shapes.

\paragraph{Deployment trade-offs}
The deployment pipeline demonstrates a clear latency--accuracy--size
trade-off. The ONNX FP32 model achieves the best throughput (11.8~FPS)
with the smallest footprint (0.5~MB), while INT8 quantization increases
the model size to 9.9~MB with a modest latency increase. For edge
deployment scenarios where model size is less constrained, the FP32 ONNX
variant offers the best performance. For INT8-accelerated hardware, the
quantized model would be preferred.

\paragraph{Limitations}
The most significant limitation of this work is the partial training
scope, which prevents definitive accuracy comparisons. Beyond this,
our approach inherits the single-class formulation of SKU-110K;
extending to fine-grained product identification (SKU-level
classification) would require a substantially larger label vocabulary.
Additionally, the mask quality of YOLACT's linear combination is
inherently limited compared to Mask R-CNN's per-ROI convolutional head.
The use of the MPS backend rather than CUDA also limited our ability to
leverage mixed-precision training and restricted batch size to 4.

\paragraph{Comparison with prior work}
Goldman~\etal~\cite{goldman2019dense} reported detection-only mAP values
on SKU-110K using RetinaNet~\cite{lin2017focal} with specialized
EM-merger post-processing, achieving mAP@0.50 of 49.2. Our Soft-NMS
approach offers a simpler alternative that does not require iterative
merging and is agnostic to the underlying detector architecture. Direct
numerical comparison with our preliminary results is not meaningful
given the partial training scope; a fully-trained variant of our pipeline
would be needed to assess competitive standing.

% ===========================================================================
% 6. CONCLUSION
% ===========================================================================
\section{Conclusion}
\label{sec:conclusion}

We presented an efficient instance segmentation pipeline for densely
packed retail scenes, combining YOLACT with a MobileNetV3-Large backbone
and Gaussian Soft-NMS post-processing. Our approach reduces backbone
parameters by 88\% compared to the standard ResNet-101 configuration
(5.4M vs.\ 44.5M backbone parameters; 10.0M total model parameters)
and achieves a 97\% reduction in FLOPs (0.22G vs.\ 7.8G). The complete
pipeline---from data preprocessing through training, evaluation, ONNX
export, and INT8 quantization---has been validated end-to-end on the
SKU-110K benchmark.

Preliminary results from training on 12\% of the data for 10 epochs
demonstrate consistent loss convergence (training loss reduced by
$3.3\times$, from 14.005 to 4.186), validating the architecture and
training pipeline. While the resulting detection accuracy (AP@0.50 =
0.08\%) reflects the limited training scope rather than the model's
potential, the monotonically decreasing loss trajectory indicates that
extended training on the full dataset would yield substantially improved
accuracy.

The deployment pipeline achieves 11.8~FPS with ONNX FP32 on CPU, a
$3.3\times$ speedup over PyTorch MPS inference, with a model footprint
under 10~MB after INT8 quantization. This demonstrates a viable path
toward edge deployment for retail analytics applications.

\paragraph{Future work}
The most immediate priority is completing full-scale training on the
entire SKU-110K training set for 80+ epochs with CUDA-equipped hardware.
Beyond this, several directions merit investigation:
(i)~knowledge distillation from a high-capacity teacher (ResNet-101
YOLACT) to further improve MobileNetV3 accuracy,
(ii)~extending the framework to multi-class SKU identification,
(iii)~incorporating temporal consistency across video frames for
real-time shelf monitoring,
(iv)~Soft-NMS $\sigma$ ablation study with a fully converged model, and
(v)~leveraging mixed-precision training on CUDA hardware to enable
larger batch sizes and faster convergence.

% ===========================================================================
% REFERENCES
% ===========================================================================
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
